Staged Abstract Interpreters

### 2018-11-12 paper/10_intro.tex
section 2/3

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
\end{figure}

>>>>

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of staging and abstraction.}
\end{figure}

### 2018-11-12 paper/21_bg_interp.tex
section 2/3

The actual semantics of the interpreter is also abstract now, such as lifting literal terms to values 
(\texttt{close}, \texttt{num}), applying a closure (\texttt{apply\_closure}), branching (\texttt{branch0},
and arithmetics (\texttt{prim\_eval}). But, just based on these abstract methods, a general interpreter 
can be defined:

@@

The implementation of the interpreter is straightforward.
It is worth to note that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively use the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

>>>>

The actual semantics and operations of the interpreter is left to be implemented later, 
such as lifting literal terms to values (\texttt{close}, \texttt{num}), 
applying a closure (\texttt{apply\_closure}), branching (\texttt{branch0}),
and arithmetics (\texttt{prim\_eval}). We can choose to implement them concrete or abstract,
unstaged or staged.
But, just based on these abstract operations, an expressive generic interpreter can be defined:

@@

The implementation of the skeleton interpreter traverses the abstract syntax tree and should be straightforward.
It is worth to note that the interpreter is written in open-recursive style -- 
it can not refer to itself directly, instead, \texttt{eval} takes an additional 
parameter \texttt{ev} referring to itself; this allows us to instrument the interpreter
from outside. Accordingly, the function \texttt{close} for lifting lambda term to closure values
is also written in this style.
We define a default implementation \texttt{fix} to close the recursion by recursively use the result 
of \texttt{fix(ev)} for \texttt{ev}. 
Finally, the top-level evaluation function \texttt{eval\_top} is defined.

### 2018-11-12 paper/22_bg_abs.tex
section 2/3

\subsection{From Interpreter to Abstract Interpreter}

Based on the generic interface we presented in Section~\ref{bg_lang}, now we describe the instantiation of an
abstract interpreter.
To keep the simplicity, we intend to use simply abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse 
but enough to setup a foundation for the staged abstract interpreter we will present later.
However, in Section \note{X}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section \note{Y}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

@@

\paragraph{Operations Implementation.}
The @get@ and @put@ operation on abstract stores slightly changed according to
the abstract semantics: @get@ uses bottom element in the @Value@ lattice if the queried
address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to a singleton set that
contains one \texttt{CloV} or \texttt{NumV} object.

\todo{conditional, should use different store}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (f: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- f) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = thn $\sqcup$ els //TODO!
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = Set(NumV())
} // to be continued
\end{lstlisting}

\paragraph{Fixpoint Iteration.}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also called co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL} and widely used in different analysis.

>>>>

\subsection{From Interpreter to Abstract Interpreter} \label{unstaged_abs}

Based on the generic interface we presented in Section~\ref{bg_lang}, 
now we describe the instantiation of an abstract interpreter.
To keep the simplicity, we intend to use simply abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse
but simple enough to setup a foundation for the staged abstract interpreter we will present later.
However, in Section~\ref{cfa}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section~\ref{cases_imp}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

@@

\paragraph{Abstract Semantics.}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

\todo{conditional, use different store?}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

For branching, @branch0@ simply joins the answer from two branches. For arithmetics, @prim_eval@ simply 
returns the top abstract number element.
It is worth to note that @apply_closure@ is where we handle the non-determinism of applications.
The non-determinism happens because of the abstract store maps addresses to sets of abstract values;
when dereferencing an address from the store, we need to explore all possible values to achieve a sound 
analysis. In the case of @apply_closure@, the first argument @fs@ may contains multiple target closures,
so we have a @for@ comprehension over all closures, apply them respectively and form a set result values @vs@;
meanwhile, every time when we evaluate into the body expression @e@ we use the latest store $\sigma$@_*@
which is updated iteratively. Thus it is a store-widened analysis.

\paragraph{Fixpoint Iteration.}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used in the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

### 2018-11-12 paper/23_bg_stage.tex
section 2/3

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as will, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and body of this function multiples @b@ five times.

\paragraph{Staged Concrete Interpreter.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.

>>>>

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as will, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\paragraph{Staged Concrete Semantics.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.

### 2018-11-12 paper/30_sai.tex
continue working on outline
intro
worked on paper
add some stuff
section 2/3

\section{Staged Abstract Interpreter}

@@

\subsection{what not to do when writing an abstract interpreter for staging}

@@

operational vs denotational

@@

\subsection{Stage Annotations}

@@

>>>>

\section{Staging Abstract Interpreters} \label{sai}

In the previous section, we have seen an unstaged abstract interpreter and a staged concrete interpreter, 
now we begin describing the implementation of their confluence -- a staged abstract interpreter. 
Unsurprisingly, the staged abstract interpreter has the same abstract semantics with the unstaged version
we presented in Section~\ref{unstaged_abs}.

@@

In this section, we present a principled approach to derive staged abstract interpreter
from its unstaged version. One guiding principle of our approach is the code of abstract semantics
and the code that optimizes should be separated. Thus it is also an advantage of doing staging 
for abstract interpreters: the designer of analysis has no need to rewrite the analysis, and 
the performance improvement comes almost for free, without any sacrifice of soundness or precision.

\paragraph{Staged Lattices.} The lattice definition we introduced in Section~\ref{unstaged_abs}
leaves space to be stage polymorphic, now we still use powerset as an example to present its 
staged version.

\begin{lstlisting}
trait RepLattice[A] extends Lattice[A, Rep]
implicit def RepSetLattice[T:Typ]: RepLattice[Set[T]] = 
  new RepLattice[Set[T]] {
    lazy val bot: Rep[Set[T]] = Set[T]()
    lazy val top: Rep[Set[T]] = throw new NotImplementedError()
    def $\sqsubseteq$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Boolean] = l1 subsetOf l2
    def $\sqcup$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Set[T]] = l1 union l2
    def $\sqcap$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Set[T]] = l1 intersect l2
  }
\end{lstlisting}

The type parameter @T:Typ@ of @RepLattice@ requires the element of sets
is also can be staged, otherwise without knowing how to stage the elements in the set
we can not stage the set as well.
The methods defined operate on type @Rep[Set[T]]@, thus the underlying implementation
such as @union@ and @intersect@ will be mapped to a node in IR graph and emitted in 
the generated code. Again, other structures such as maps and tuples are also implemented
in a similar way.

\paragraph{Staged Abstract Semantics.}

\begin{lstlisting}
trait RepAbsInterpOps extends Abstract with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident, Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr, Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): 
    Rep[Env] = $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = 
    $\sigma$.getOrElse(a, RepLattice[Value].bot)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): 
    Rep[Store] = $\sigma$ + (a -> RepLattice[Value].$\sqcup$(v, get($\sigma$, a)))
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = Addr(x)
  def num(i: Lit): Rep[Value] = Set(NumV())
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = $\lambda$
    val f: Rep[(Value, Store)]=>Rep[(Value,Store)] = 
      (as: Rep[(Value,Store)]) => {
        val args = as._1; val $\sigma$ = as._2; val $\alpha$ = alloc($\sigma$, x)
        ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, args))
      }
    Set[AbsValue](CompiledClo(fun(f)))
  }
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans =
    thn $\sqcup$ els
  def prim_eval(op: Symbol, 
                v1: Rep[Value], v2: Rep[Value]): Rep[Value] = 
    Set(NumV())
  ...
}
\end{lstlisting}

\paragraph{Staged Fixpoint Iteration.} staged fixed point iteration with memoization

\paragraph{Specialized Data Structures.}

\subsection{Discussion}
what not to do when writing an abstract interpreter for staging?
what kind of abstract interpreter is friendly for staging/specialization?

Stage an abstract abstract machine yields another small-step abstract machine

@@

direct-style vs CPS

Operational vs denotational

@@

\cite{10.1007/3-540-61580-6_11}

@@

\subsection{Lattice-theoretic Interpretation}
